Scalar Feedback based Joint Time-Frequency Precoder interpolation for MIMO-OFDM Systems

### 2019-10-17 paper/main.tex
Add text

% \noindent We consider problem of quantization and interpolation of
% time and frequency varying precoding matrices in wireless MIMO
\begin{abstract}
In a limited feedback MIMO channel, the performance of the channel can improve significantly if the transmitter knows the channel state information(CSI). The receiver knows the channel information, and it feeds back to the transmitter. However, it is not possible to feedback complete information for a limited-feedback OFDM channel since the channel information for multiple subcarriers consumes a large amount of data. Therefore to resolve this problem, the orthonormality structure of the precoder matrix is exploited to make its one-to-one mapping to a minimal number of scalar parameters. The Precoder matrix is decomposed using Givens Rotation. For an i.i.d. flat-fading Rayleigh channel, these parameters are independent, which makes its quantization and interpolation easy. We propose an efficient method for quantization of the channel state information at the receiver by using adaptive delta modulation for successive time interval. We also propose a method for joint time-frequency interpolation of the channel information at the transmitter. 153-words
\end{abstract}

@@

The problem of effective quantization and feedback of precoders for MIMO-OFDM has attracted significant attention in recent years. Many approaches involve codebook design over the Grassmann manifold~\cite{mondal2007quantization,schwarz2013adaptive,5946308} or the Stiefel Manifold \cite{6891198,Gupt1905:Predictive}, wherein the manifold structure of the precoders is used to obtain tangents for predictive quantization and interpolation across time and frequency. While these methods have been shown to be very effective for precoder feedback,  they involve operations over higher dimensional manifolds. In particular, optimization and interpolation is complicated, especially for manifolds where geodesic paths for interpolation may be difficult to obtain, especially for joint time-frequency predictive coding. An alternate approach is to parametrize the precoding matrices using scalar parameters. The space of unitary matrices parameterized into independent scalar parameters that correspond to Givens rotations and Household Transformations of the unitary precoding matrices \cite{4114278,4556174}. These approaches have been combined with adaptive delta modulation effectively track precoding matrices both in the time~\cite{4114278} and frequency domains~\cite{4556174}, albeit separately.

@@

\section{Simulation and discussion}
\label{section3}
We present some simulations of the scalar feedback based joint
time-frequency predictive quantization outlined in the previous
section. In particular, we analyze the performance of the proposed
quantization and interpolation method for time-varying MIMO-OFDM
systems. The channel is simulated using the typical Jakes Model for
both the ITU Vehicular-A channel and ITU Pedestrian channel power
delay profiles CITE. The BER achieved by the proposed predictive
quantization based precoder is compared against the ideal precoder for
100 channel evolutions in each run and averaged. The normalised
Doppler is assumed to be $f_dT_s = 3.5\times10^{-4}$, with symbol time
period $T_s = 5\times10^{-8} s$. For both the Vehicular and Pedestrian
channel profiles, we consider $N_T=4$ and $N_R=2$. Therefore, the
total number of parameters that determine the precoder are
$N_{R}(2N_{T} - N_{R}) = 12$, with 5$\theta_l$'s and
7$\phi_i$'s. For the Vehicular channel profile, we assume that the
number of subcarriers is 1024, while for the Pedestrian channel
profile it is 64. We take the spacing between fed back subcarrier
indices $p$ to be $33$ and $9$ for the Vehicular
and Pedestrian channel profiles respectively (i.e. for the Vehicular
profile, the subcarriers indices fed back would be $0, 33, 66, \ldots
1023$ and for the Pedestrian, $0, 9, 18, \ldots 63$).

For the first and the second simulation time instances, since the
precoder is not available at the transmitter, we use 10 bits for
quantizing each parameter for effective representation of the
precoder. Therefore, the number of bits for each subcarrier =
$10\times12$ = 120. Quantization is performed as given in
Section~\ref{section2}. We note that this is not significant, since,
amortized over long durations, this number becomes small. Subsequent
quantization takes place in the time-domain using one bit for each
parameter. Since the bit budget is small, if the channel varies
sufficiently slowly, either the $\theta_l$ or the $\phi_i$ parameters
are fed back alternately, while retaining the unsent parameters as
being to their previous values, as shows in
Fig.~\ref{fig:adpm-fig}. Here, $\Theta$ is the the collection of all
$\theta_l$s and $\Phi$ is the collection of all $\phi_i$s. This brings
down the bit budget to an average of 5 bits(the length of $\Theta$)
and 7 bits (the length of $\Phi$) over alternate feedback instances,
thereby resulting in an average feedback budget of 5 bits.

To enable adaptive tracking of the channel parameters, we start with
an initialization of the step size for each parameter, as discussed in
Equation~\ref{delta_eqn}, with $M = 1.4$ (the choice is motivated by
the results in~\cite{5671092}), and track parameters with the temporal
evolution of the channel, and we find that this choice effectively
tracks the channel parameters.

% Performance Measurements:
% Since the Precoding matrices fundamentally lie on the Stiefel manifold, therefore, we are using the chordal distance parameter to measure the effectiveness of the quantization method used by us. Later we are comparing the BER rates with the ideal completely fed back subcarriers.

@@

% %talk about this in the simulations, that would be more relevant
% Since $\Delta_{n} = M\Delta_{n-1} or \Delta_{n-1}/m $, we found empirically that $m=2.5*M$ i.e. decay

\begin{figure}
\includegraphics[width=0.5\textwidth]{images/pedestrian.pdf}
\caption{BER vs SNR for pedestrian channel}
\label{ber_overview}
\vspace{-5pt}
\end{figure}

Since the Precoding matrices fundamentally lie on the Stiefel
manifold, therefore, we are using the chordal distance parameter to
measure the effectiveness of the quantization method used. Later we
are comparing the BER rates with the completely fed back
subcarriers. Or simulations and discussions part 2 criteria could be
used. Done similar work as in \cite{4114278} but along with that we
have joint interpolated in time and frequency at the transmitter using
the fundamental scalar parameters obtained. To lower the number of
bits we have used the method of interpolation and have achieved
significant improvement over predictive quantization method in
\cite{6891198} and also in \cite{Gupt1905:Predictive} which tried to
use joint interpolation over the tangent space in the Stiefel
Manifold. In fact, we use joint interpolation of the scalar parameters
which are easy to use and give better quantization than any of the
above methods.

The advantage they had over the number of bits due to the 6 bit codebook in \cite{6891198,Gupt1905:Predictive} is also achieved by us by using smart interpolation techniques, i.e. by dropping the feedback bits in alternating time instances as it is still going to follow the scalar parameters without much difference.

The problem of values hopping between $\pi$ and $-\pi$ is also tackled by wrapping the values around which we can show works nicely. The only drawback is that while initializing the parameter, they go out of their actual range and therefore using a uniform quantizer over the prescribed range does not work. But since most of the values lie between the given range we can use most of the initialization bits quantizing the parameters uniformly between the [-$\pi$ , $\pi$] and using smaller amount of bits between the range outside that which could go up to $-3\pi$ and $3\pi$, i.e. we are covering 300% more range outside the given area and that works mostly fine for the rest of the values.

Using the idea of joint interpolation given in the other paper.

\begin{figure}
\includegraphics[width=0.5\textwidth]{images/vehicular_ber.pdf}
\caption{BER vehicular plot comparison between our method and \cite{Gupt1905:Predictive}}
\label{ber_overview}
\vspace{-5pt}
\end{figure}
%Qtisn error with 1000 chan evols
%Qtisn error with 100 chan evols
%BER
%Achievable rat
\section{Conclusions and Future Work}
\label{section4}

We could reduce more by seeing the effect of sigma values (power distribution). We can save the bits by not sending for those which have very low sigma values. This could result in a very robust method.

\vspace{-4pt}

>>>>

% \noindent We consider problem of quantization and interpolation of
% time and frequency varying precoding matrices in wireless MIMO
\begin{abstract}
% In a limited feedback MIMO channel, the performance of the channel can improve significantly if the transmitter knows the channel state information(CSI). The receiver knows the channel information, and it feeds back to the transmitter. However, it is not possible to feedback complete information for a limited-feedback OFDM channel since the channel information for multiple subcarriers consumes a large amount of data. Therefore to resolve this problem, the orthonormality structure of the precoder matrix is exploited to make its one-to-one mapping to a minimal number of scalar parameters. The Precoder matrix is decomposed using Givens Rotation. For an i.i.d. flat-fading Rayleigh channel, these parameters are independent, which makes its quantization and interpolation easy. We propose an efficient method for quantization of the channel state information at the receiver by using adaptive delta modulation for successive time interval. We also propose a method for joint time-frequency interpolation of the channel information at the transmitter. 153-words

  In MIMO-OFDM systems, knowledge of the precoding matrix enables
  effective resource allocation for high rates and low BER. However,
  feeding back precoders for all subcarriers introduces large
  overheads. Past work has shown that the orthonormal structure of the
  precoder enables effective parameterization and quantization using
  Givens rotations and Householder transformations. In addition, the
  use of temporal and frequency correlations can further reduce the
  feedback requirements.  We present a framework that utilizes
  adaptive feedback with joint time-frequency prediction for precoder
  construction at the transmitter.  Simulations reveal that the
  proposed method achieves low BERs for various channel profiles
  compared to other approaches.
\end{abstract}

@@

The problem of effective quantization and feedback of precoders for
MIMO-OFDM has attracted significant attention in recent years. Many
approaches involve codebook design over the Grassmann
manifold~\cite{mondal2007quantization,schwarz2013adaptive,5946308},
the Stiefel manifold \cite{6891198,Gupt1905:Predictive} and the Flag
manifold~\cite{pitaval2013codebooks}, wherein the manifold structure
of the precoders is used to obtain tangents for predictive
quantization and interpolation across time and frequency. While these
methods have been shown to be very effective for precoder feedback,
they involve operations over higher dimensional manifolds. In
particular, optimization and interpolation is complicated, especially
for manifolds where geodesic paths for interpolation may be difficult
to obtain, especially for joint time-frequency predictive coding. An
alternate approach is to parametrize the precoding matrices using
scalar parameters. The space of unitary matrices parameterized into
independent scalar parameters that correspond to Givens rotations and
Household Transformations of the unitary precoding matrices
\cite{4114278,4556174}. These approaches have been combined with
adaptive delta modulation effectively track precoding matrices both in
the time~\cite{4114278} and frequency domains~\cite{4556174}, albeit
separately.

@@

\section{Simulation and discussion}
\label{section3}
We now present simulations of the scalar feedback based joint
time-frequency predictive quantization outlined in the previous
section. In particular, we analyze the performance of the proposed
quantization and interpolation method for time-varying MIMO-OFDM
systems. The channel is simulated using the typical Jakes Model for
both the ITU Vehicular-A channel and ITU Pedestrian-A channel power
delay profiles~\cite{recommendation1997guidelines}. The BER achieved by the proposed predictive
quantization based precoder is compared against the ideal precoder for
100 channel evolutions in each run and averaged. The normalised
Doppler is assumed to be $f_dT_s = 3.5\times10^{-4}$, with symbol time
period $T_s = 5\times10^{-8} s$. For both the Vehicular and Pedestrian
channel profiles, we consider $N_T=4$ and $N_R=2$. Therefore, the
total number of parameters that determine the precoder are
$N_{R}(2N_{T} - N_{R}) = 12$, with 5$\theta_l$s and
7$\phi_i$s. For the Vehicular channel profile, we assume that the
number of subcarriers is 1024, while for the Pedestrian channel
profile it is 64. We take the spacing between fed back subcarrier
indices $p$ to be $33$ and $9$ for the Vehicular
and Pedestrian channel profiles respectively (i.e. for the Vehicular
profile, the subcarriers indices fed back would be $0, 33, 66, \ldots
1023$ and for the Pedestrian, $0, 9, 18, \ldots 63$).

For the first and the second simulation time instances, since the
precoder is not available at the transmitter, we use 2 bits for
initializing each parameter for effective representation of the
precoder. Therefore, the number of bits for each subcarrier =
$10\times 2$ = 20 (we note that $\phi_0 = \phi_4 = 0$ due to the
non-uniqueness of the SVD). Quantization is performed as given in
Section~\ref{section2}. We note that this is not significant, since,
amortized over long durations, this number becomes small. Subsequent
quantization takes place in the time-domain using one bit for each
parameter. Since the bit budget is small, if the channel varies
sufficiently slowly, either the $\theta_l$ or the $\phi_i$ parameters
are fed back alternately, while retaining the unsent parameters as
being to their previous values, as shown in
Fig.~\ref{fig:adpm-fig}. Here, $\Theta$ is the the collection of all
$\theta_l$s and $\Phi$ is the collection of all $\phi_i$s. This brings
down the bit budget to an average of 5 bits(the length of $\Theta$)
and 7 bits (the length of $\Phi$) over alternate feedback instances,
thereby resulting in an average feedback budget of 5 bits.

To enable adaptive tracking of the channel parameters, we start with
an initialization of the step size for each parameter, as discussed in
Equation~\ref{delta_eqn}. The value of $M$ was chosen to be $1.4$, and
can be adjusted appropriately to track the temporal evolution of
parameters for different Doppler frequencies.

Fig.~\ref{fig:error_decay}, shows the rate at which error diminishes
with time for the proposed approach and direct quantization on the
Stiefel manifold using codebooks described
in~\cite{Gupt1905:Predictive}. We see that, since the adaptation of
independent scalar parameters causes the error to decay rapidly, the
error is lower and the convergence is faster.
\begin{figure}
\includegraphics[width=0.5\textwidth]{images/stief_veh.pdf}
\caption{\label{fig:error_decay}A comparison of the $4\times 2$ precoder quantization error vs. time instant for
  the proposed approach (labelled ``TODO'') and direct quantization on
the Stiefel manifold for the Vehicular channel profile.}
\end{figure}

TODO: Agrim and Mohit discuss and then fill this part.
\begin{figure}
\includegraphics[width=0.5\textwidth]{images/ber-pedestrian-35e-4}
\caption{BER vs. SNR for QPSK transmission over the $4\times 2$ MIMO
  channel with the ITU Pedestrian-A channel profile for $f_dT_s = 3.5\times 10^{-4}$.}
\label{fig:ber_ped}
\end{figure}

Fig.~\ref{fig:ber_ped} compares the BER obtained using the proposed
quantization method with that obtained using exact precoders. We can
observe that the performance of the proposed method is very close to
the performance with exact precoders. This can be attributed to the
fact that even the 1 bit update to the scalar parameters is able to
approximate the precoder very accurately. Since the channel variation
across subcarriers is gradual, owing to relatively small delay spread
of the Pedestrian channel profile, linear interpolation and prediction
yields good performance. Moreover, this method outperforms the Stiefel
manifold based approach to the fact that the adaptation is much more
accurate for scalar parameters than the tangent space based adaptation
used for the Stiefel manifold.

Fig.~\ref{fig:ber_veh} shows a similar comparison for the Vehicular
channel profile. The larger delay spread of the channel necessitates
the use of a larger number of subcarriers (1024) than the Pedestrian
channel (64). The larger number of subcarriers implies that the
predictive quantization and interpolation needs to be performed across
a larger frequency range, thereby reducing the accuracy. Even in this
case, we observe that the scalar parameter based approach outperforms
manifold based quantization by about 5 dB.

\begin{figure}
\includegraphics[width=0.5\textwidth]{images/ber-vehicular-35e-4}
\caption{BER vs. SNR for QPSK transmission over the $4\times 2$ MIMO
  channel with the ITU Vehicular-A channel profile for $f_dT_s = 3.5\times 10^{-4}$.}
\label{fig:ber_veh}
\end{figure}

% Since the Precoding matrices fundamentally lie on the Stiefel
% manifold, therefore, we are using the chordal distance parameter to
% measure the effectiveness of the quantization method used. Later we
% are comparing the BER rates with the completely fed back
% subcarriers. Or simulations and discussions part 2 criteria could be
% used. Done similar work as in \cite{4114278} but along with that we
% have joint interpolated in time and frequency at the transmitter using
% the fundamental scalar parameters obtained. To lower the number of
% bits we have used the method of interpolation and have achieved
% significant improvement over predictive quantization method in
% \cite{6891198} and also in \cite{Gupt1905:Predictive} which tried to
% use joint interpolation over the tangent space in the Stiefel
% Manifold. In fact, we use joint interpolation of the scalar parameters
% which are easy to use and give better quantization than any of the
% above methods.

% The advantage they had over the number of bits due to the 6 bit codebook in \cite{6891198,Gupt1905:Predictive} is also achieved by us by using smart interpolation techniques, i.e. by dropping the feedback bits in alternating time instances as it is still going to follow the scalar parameters without much difference.

% The problem of values hopping between $\pi$ and $-\pi$ is also tackled by wrapping the values around which we can show works nicely. The only drawback is that while initializing the parameter, they go out of their actual range and therefore using a uniform quantizer over the prescribed range does not work. But since most of the values lie between the given range we can use most of the initialization bits quantizing the parameters uniformly between the [-$\pi$ , $\pi$] and using smaller amount of bits between the range outside that which could go up to $-3\pi$ and $3\pi$, i.e. we are covering 300% more range outside the given area and that works mostly fine for the rest of the values.

@@

\section{Conclusion}
\label{section4}
We have presented a predictive quantization based precoder
reconstruction of a MIMO-OFDM system that employs scalar quantization
based feedback of precoder parameters. By quantizing and adapting the
Givens rotation and Householder trasnsformation parameters of the
precoder, the receiver is able to send minimal feedback to the
transmitter and yet enable effective precoder adaptation. Exploiting
the temporal and frequency correlations jointly enhances precoder
accuracy significantly. Simulations reveal that the proposed approach
provides faster convergence and smaller errors in the precoder when
compared to approaches that involve direct quantization on manifold
tangent spaces. Future work would involve tuning the adaptation rate
of the scalar parameters and extending this to TODO.
